---
title: "STA openness (liberalism and intellect)"
format: gfm
---

This repository accompanies the paper *Perceived Liberalism and Intellect Cannot Be Reduced to a Single Openness Dimension: A State-Trace Experiment*

## Repository contents

- [`materials.md`](materials.md) documents the vignette procedure, all six
  experimental conditions, the exact German item wording, response scales,
  outcome measures, manipulation checks, and attention checks.
- `between.csv` and `within.csv` are the analysis-ready datasets used below.
- `Main_Study/Data` contains the main exported dataset and its data
  documentation.
- `Main_Study/Scripts` contains the detailed original analysis workflow.
- `Pilot_Study` contains the corresponding pilot-study data, documentation,
  and scripts.

The repository is connected to the article's anonymous OSF project for peer
review. The OSF project will be made public for publication.

Note that this repository corresponds to the *article* version of the analysis. Compared to the original thesis scripts, the code has been substantially shortened and simplified.

If you require more detailed data handling or additional analyses, please refer to the `Main_Study` folder. Be aware, however, that the code there is considerably more difficult to follow. In particular, the original implementation relies on the *STACMR-R* scripts, which must be installed manually because they are not properly packaged for R.

In the present file, we instead use a wrapper package called *stacmr*, which enables a more streamlined and reproducible workflow. If your goal is simply to reproduce the reported results or to conduct additional analyses using the main dataset, you can proceed with the code below.

## Required packages

We begin by loading the required libraries. Throughout this project, we use **librarian** as a package manager. If you do not have it installed, please install it first or substitute it with your preferred alternative.

```{r}
# A recent JRP production article uses Charis SIL for body text. The font and its SIL
# Open Font License are vendored in this repository so figure export does not
# depend on a machine-wide font installation.
figure_font <- "Charis SIL"
font_config <- normalizePath("fonts/fonts.conf", mustWork = TRUE)
Sys.setenv(FONTCONFIG_FILE = font_config)
font_match <- system2("fc-match", shQuote(figure_font), stdout = TRUE)
if (!length(font_match) || !grepl("CharisSIL-Regular.ttf", font_match[[1]], fixed = TRUE)) {
  stop("The vendored Charis SIL font could not be resolved through fontconfig.")
}

librarian::shelf(car, readr, dplyr, tidyr, effectsize, psych, lsr,
devtools, monotonicity/stacmr, tidyverse, simpleCache)
setCacheDir("cache")
```

## Between analysis

Load the data:

```{r}
d <- read_csv("between.csv",
              col_type = list(iv_se = "f", iv_ai = "f", iv_comb = "f"))
# correct factor ordering
d$iv_se <- forcats::fct_relevel(d$iv_se, c("low", "high"))
d$iv_ai <- forcats::fct_relevel(d$iv_ai, c("low", "medium", "high"))
```

### Manipulation check with C05SE and O98AI:

SE: Self-Efficacy
AI: Artistic Interest

The items **C05SE** and **O98AI** served as manipulation checks and were not included in the vignette itself. If the vignette successfully induces the intended manipulation, these items should show a strong correlation with the manipulation variable.

```{r}
cor.test(d$C05SE, as.numeric(d$iv_se))
cor.test(d$O98AI, as.numeric(d$iv_ai), method = "s")
```

### STA

make long format:

```{r}
between_long <- as.data.frame(pivot_longer(d,
  cols = c("LibSum", "IntSum"),
  names_to = "DV",
  values_to = "Sum"
)) 
between_long$DV <- as.factor(between_long$DV)
```

```{r}
stats <- sta_stats(data = between_long, col_value = "Sum", 
                   col_participant = "id",
                   col_dv = "DV", col_between = "iv_comb")
stats
```
We would usually specify a partial order at this stage. However, since the partial
order already exhibits a perfect fit, this step can be omitted.

```{r}
set.seed(1)
simpleCache("res1", {cmr(between_long, col_value = "Sum", 
                         col_participant = "id",
                         col_dv = "DV", col_between = "iv_comb", nsample=1e5)})
res1

res1$estimate
```

### ST Plot

Although the *stacmr* package provides a dedicated function for this purpose, it
does not produce a publication-ready plot. We therefore extract the relevant
data and construct the ST plot manually.

```{r}
# Data for the Plot
d_agg <- data.frame(x_mean = stats$LibSum$means, y_mean = stats$IntSum$means,
                    x_se = sqrt(diag(stats$LibSum$cov) / diag(stats$LibSum$n)),
                    y_se = sqrt(diag(stats$IntSum$cov) / diag(stats$IntSum$n)),
                    iv_comb = stats$IntSum$conditions)

d_agg <- d_agg %>% mutate(x_min = x_mean - x_se,  # Define lower and upper limits for error bars
  x_max = x_mean + x_se,
  y_min = y_mean - y_se,
  y_max = y_mean + y_se)

d_agg <- d_agg %>%
  separate(iv_comb, into = c("Self-Efficacy", "Artistic Interest"))
# correct factor ordering
d_agg$`Self-Efficacy` <- forcats::fct_relevel(d_agg$`Self-Efficacy`, c("low", "high"))
d_agg$`Artistic Interest` <- forcats::fct_relevel(d_agg$`Artistic Interest`, c("low", "medium", "high"))

d_cmr <- as.data.frame(res1$estimate)

# ggplot
p1 <- ggplot(d_agg, aes(x_mean, y_mean)) +
  geom_point(data = d_cmr, aes(LibSum, IntSum), color = "grey60",
             shape = 5, size = 3) +  # Add predicted Points (Regression)
  geom_line(data = d_cmr, aes(LibSum, IntSum), color = "grey60", linetype = "dashed") +  # grey curve for Regression model
  geom_errorbar(aes(ymin = y_min, ymax = y_max), color = "grey60", width = 0.2) +  # y-Axis Error bars
  geom_errorbarh(aes(xmin = x_min, xmax = x_max), color = "grey60", height = 0.2) +  # Add x-Axis Error Bars
  geom_point(aes(shape = `Artistic Interest`, fill = `Self-Efficacy`),
             color = "black", size = 3) +  # open low SE, filled high SE
  scale_shape_manual(values = c(21, 24, 22)) +  # fillable circle, triangle, square
  scale_fill_manual(values = c(low = "white", high = "black")) +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +

  labs(
    x = "Liberalism",
    y = "Intellect",
  ) +
  # xlim(4, 20) +
  # ylim(4, 20) +
  
  theme_bw(base_family = figure_font) +
  theme(
    text = element_text(family = figure_font),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.grid = element_blank(),
    legend.background = element_blank(),
    legend.key = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.box = "horizontal",
    legend.position = c(0.02, 0.98),   # x, y in NPC coordinates (0 = left/bottom, 1 = right/top)
    legend.justification = c("left", "top")
  ) +
  coord_fixed(ratio = 1) + 
  annotate("text", 12.7, 11.5, label = "dashed line shows\nbest isotonic model",
           hjust = 0, family = figure_font)
p1
```

## Within analysis

```{r}
d <- read_csv("within.csv",
              col_type = list(iv_se = "f", iv_ai = "f", iv_comb = "f",
                              dv = "f"))
# correct factor ordering
d$iv_se <- forcats::fct_relevel(d$iv_se, c("low", "high"))
d$iv_ai <- forcats::fct_relevel(d$iv_ai, c("low", "medium", "high"))
within = as.data.frame(d)
```


```{r}
stats <- sta_stats(data = within, col_value = "sum", 
                   col_participant = "id",
                   col_dv = "dv", col_within = "iv_comb")
stats
```

We would usually specify a partial order at this stage. However, since the partial
order already exhibits a perfect fit, this step can be omitted.

```{r}
set.seed(1)
simpleCache("res2", {cmr(within, col_value = "sum", 
                         col_participant = "id",
                         col_dv = "dv", col_within = "iv_comb", nsample = 1e5)})
res2
res2$estimate
```

### ST Plot

Although the *stacmr* package provides a dedicated function for this purpose, it
does not produce a publication-ready plot. We therefore extract the relevant
data and construct the ST plot manually.

Note that in the thesis, standard errors were not calculated
correctly and were consequently overestimated. The approach used below follows
exactly the same procedure as implemented in the *stacmr* package.

```{r}
# Data for the Plot
d_agg <- data.frame(x_mean = stats$LibSum$means, y_mean = stats$IntSum$means,
                    x_se = sqrt(diag(stats$LibSum$cov) / diag(stats$LibSum$n)),
                    y_se = sqrt(diag(stats$IntSum$cov) / diag(stats$IntSum$n)),
                    iv_comb = names(stats$IntSum$conditions))

d_agg <- d_agg %>% mutate(x_min = x_mean - x_se,  # Define lower and upper limits for error bars
  x_max = x_mean + x_se,
  y_min = y_mean - y_se,
  y_max = y_mean + y_se)
d_agg

d_agg$iv_comb = rownames(d_agg)

d_agg <- d_agg %>%
  separate(iv_comb, into = c("Self-Efficacy", "Artistic Interest"))
# correct factor ordering
d_agg$`Self-Efficacy` <- forcats::fct_relevel(d_agg$`Self-Efficacy`, c("low", "high"))
d_agg$`Artistic Interest` <- forcats::fct_relevel(d_agg$`Artistic Interest`, c("low", "medium", "high"))

d_cmr <- as.data.frame(res2$estimate)

# ggplot
p2 <- ggplot(d_agg, aes(x_mean, y_mean)) +
  geom_point(data = d_cmr, aes(LibSum, IntSum), color = "grey60",
             shape = 5, size = 3) +  # Add predicted Points (Regression)
  geom_line(data = d_cmr, aes(LibSum, IntSum), color = "grey60", linetype = "dashed") +  # grey curve for Regression model
  geom_errorbar(aes(ymin = y_min, ymax = y_max), color = "grey60", width = 0.2) +  # y-Axis Error bars
  geom_errorbarh(aes(xmin = x_min, xmax = x_max), color = "grey60", height = 0.2) +  # Add x-Axis Error Bars
  geom_point(aes(shape = `Artistic Interest`, fill = `Self-Efficacy`),
             color = "black", size = 3) +  # open low SE, filled high SE
  scale_shape_manual(values = c(21, 24, 22)) +  # fillable circle, triangle, square
  scale_fill_manual(values = c(low = "white", high = "black")) +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  labs(
    x = "Liberalism",
    y = "Intellect",
  ) +
   xlim(7, 17.5) +
   ylim(7, 17.5) +
  
  theme_bw(base_family = figure_font) +
  theme(
    text = element_text(family = figure_font),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.grid = element_blank(),
    legend.background = element_blank(),
    legend.key = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.box = "horizontal",
    legend.position = c(0.02, 0.98),   # x, y in NPC coordinates (0 = left/bottom, 1 = right/top)
    legend.justification = c("left", "top")
  ) +
  coord_fixed(ratio = 1) + 
  annotate("text", 11.4, 12.25, label = "dashed line shows\nbest isotonic model",
           hjust = 0, family = figure_font)
p2
```


```{r}
#| eval: true
#| include: false
cairo_pdf("between.pdf", width = 6, height = 6, family = figure_font)
print(p1)
dev.off()
cairo_pdf("within.pdf", width = 6, height = 6, family = figure_font)
print(p2)
dev.off()

cairo_ps("between.eps", width = 6, height = 6, family = figure_font,
         onefile = FALSE, fallback_resolution = 600)
print(p1)
dev.off()
cairo_ps("within.eps", width = 6, height = 6, family = figure_font,
         onefile = FALSE, fallback_resolution = 600)
print(p2)
dev.off()

manuscript_plot_dir <- normalizePath("../latex/plots", mustWork = TRUE)
figure_files <- c("between.pdf", "within.pdf", "between.eps", "within.eps")
copied <- file.copy(figure_files, manuscript_plot_dir, overwrite = TRUE)
stopifnot(all(copied))
```
