---
title: "STA openness (liberalism and intellect)"
format: gfm
---

This repository contains the anonymized data, study materials, and R analysis
code accompanying the paper *Perceived Liberalism and Intellect Cannot Be
Reduced to a Single Openness Dimension: A State-Trace Experiment*.

## Citation

Until the article receives its DOI and final bibliographic details, please cite
it as:

> Titz, J., & Gailer, B. (in press). Perceived liberalism and intellect cannot
> be reduced to a single openness dimension: A state-trace experiment.
> *Journal of Research in Personality*.

Machine-readable citation metadata are provided in
[`CITATION.cff`](CITATION.cff). Versioned releases of this repository are
archived on Zenodo.

## Repository contents

- [`materials.md`](materials.md) documents the vignette procedure, all six
  experimental conditions, the exact German item wording, response scales,
  outcome measures, manipulation checks, and attention checks.
- `between.csv` and `within.csv` are the analysis-ready datasets used below.
- `Main_Study/Data` contains the main exported dataset and its data
  documentation.
- `Main_Study/Scripts` contains the detailed original analysis workflow.
- `Pilot_Study` contains the corresponding pilot-study data, documentation,
  and scripts.

## Licensing

Licensing varies by content: the R code is available under GPL-3.0-or-later;
original data, documentation, and study materials are available under CC BY
4.0; and the underlying IPIP items and scales are in the public domain. See
[`LICENSE.md`](LICENSE.md) for the precise scope and third-party notices.

## Reproducing the article analyses

This README presents the streamlined analyses reported in the article. Compared
with the original thesis scripts, the code has been substantially shortened and
simplified.

For detailed data handling and additional analyses, see the `Main_Study`
folder. Those original scripts are more complex and rely on the *STACMR-R*
scripts, which require manual installation.

The workflow below instead uses the *stacmr* wrapper package. It is the
recommended starting point for reproducing the reported results or conducting
additional analyses with the article dataset.

### Required software and packages

The analysis requires a current R installation. The code below uses
**librarian** to install any missing packages on first use and then load them.
Install **librarian** once if it is not already available:

```r
install.packages("librarian")
```

The *stacmr* package depends on **rJava**, so Java must also be installed and
configured for the same architecture as R. If the package installation fails
at `rJava`, configure Java for R and then rerun the package-loading chunk.

Then load the packages required by the streamlined workflow:

```{r}
librarian::shelf(
  readr, dplyr, tidyr, forcats, ggplot2, simpleCache,
  monotonicity/stacmr
)
setCacheDir("cache")
```

### Between-subjects analysis

Load the data:

```{r}
d <- read_csv("between.csv",
              col_type = list(iv_se = "f", iv_ai = "f", iv_comb = "f"))
# correct factor ordering
d$iv_se <- forcats::fct_relevel(d$iv_se, c("low", "high"))
d$iv_ai <- forcats::fct_relevel(d$iv_ai, c("low", "medium", "high"))
```

#### Manipulation checks

- **SE:** self-efficacy
- **AI:** artistic interest

The items **C05SE** and **O98AI** served as manipulation checks and were not included in the vignette itself. If the vignette successfully induces the intended manipulation, these items should show a strong correlation with the manipulation variable.

```{r}
cor.test(d$C05SE, as.numeric(d$iv_se))
cor.test(d$O98AI, as.numeric(d$iv_ai), method = "s")
```

#### State-trace analysis

Reshape the data from wide to long format:

```{r}
between_long <- as.data.frame(pivot_longer(d,
  cols = c("LibSum", "IntSum"),
  names_to = "DV",
  values_to = "Sum"
)) 
between_long$DV <- as.factor(between_long$DV)
```

```{r}
stats <- sta_stats(data = between_long, col_value = "Sum", 
                   col_participant = "id",
                   col_dv = "DV", col_between = "iv_comb")
stats
```
We would usually specify a partial order at this stage. However, since the partial
order already exhibits a perfect fit, this step can be omitted.

```{r}
set.seed(1)
simpleCache("res1", {cmr(between_long, col_value = "Sum", 
                         col_participant = "id",
                         col_dv = "DV", col_between = "iv_comb", nsample=1e5)})
res1

res1$estimate
```

#### State-trace plot

Although the *stacmr* package provides a dedicated function for this purpose, it
does not produce a publication-ready plot. We therefore extract the relevant
data and construct the ST plot manually.

```{r}
# Data for the Plot
d_agg <- data.frame(x_mean = stats$LibSum$means, y_mean = stats$IntSum$means,
                    x_se = sqrt(diag(stats$LibSum$cov) / diag(stats$LibSum$n)),
                    y_se = sqrt(diag(stats$IntSum$cov) / diag(stats$IntSum$n)),
                    iv_comb = stats$IntSum$conditions)

d_agg <- d_agg %>% mutate(x_min = x_mean - x_se,  # Define lower and upper limits for error bars
  x_max = x_mean + x_se,
  y_min = y_mean - y_se,
  y_max = y_mean + y_se)

d_agg <- d_agg %>%
  separate(iv_comb, into = c("Self-Efficacy", "Artistic Interest"))
# correct factor ordering
d_agg$`Self-Efficacy` <- forcats::fct_relevel(d_agg$`Self-Efficacy`, c("low", "high"))
d_agg$`Artistic Interest` <- forcats::fct_relevel(d_agg$`Artistic Interest`, c("low", "medium", "high"))

d_cmr <- as.data.frame(res1$estimate)

# ggplot
p1 <- ggplot(d_agg, aes(x_mean, y_mean)) +
  geom_point(data = d_cmr, aes(LibSum, IntSum), color = "grey60",
             shape = 5, size = 3) +  # Add predicted Points (Regression)
  geom_line(data = d_cmr, aes(LibSum, IntSum), color = "grey60", linetype = "dashed") +  # grey curve for Regression model
  geom_errorbar(aes(ymin = y_min, ymax = y_max), color = "grey60", width = 0.2) +  # y-Axis Error bars
  geom_errorbar(aes(xmin = x_min, xmax = x_max), color = "grey60",
                orientation = "y", width = 0.2) +  # Add x-Axis Error Bars
  geom_point(aes(shape = `Artistic Interest`, fill = `Self-Efficacy`),
             color = "black", size = 3) +  # open low SE, filled high SE
  scale_shape_manual(values = c(21, 24, 22)) +  # fillable circle, triangle, square
  scale_fill_manual(values = c(low = "white", high = "black")) +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +

  labs(
    x = "Liberalism",
    y = "Intellect",
  ) +
  # xlim(4, 20) +
  # ylim(4, 20) +
  
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.grid = element_blank(),
    legend.background = element_blank(),
    legend.key = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.box = "horizontal",
    legend.position = c(0.02, 0.98),   # x, y in NPC coordinates (0 = left/bottom, 1 = right/top)
    legend.justification = c("left", "top")
  ) +
  coord_fixed(ratio = 1) + 
  annotate("text", 12.7, 11.5, label = "dashed line shows\nbest isotonic model",
           hjust = 0)
p1
```

### Within-subjects analysis

```{r}
d <- read_csv("within.csv",
              col_type = list(iv_se = "f", iv_ai = "f", iv_comb = "f",
                              dv = "f"))
# correct factor ordering
d$iv_se <- forcats::fct_relevel(d$iv_se, c("low", "high"))
d$iv_ai <- forcats::fct_relevel(d$iv_ai, c("low", "medium", "high"))
within <- as.data.frame(d)
```


```{r}
stats <- sta_stats(data = within, col_value = "sum", 
                   col_participant = "id",
                   col_dv = "dv", col_within = "iv_comb")
stats
```

We would usually specify a partial order at this stage. However, since the partial
order already exhibits a perfect fit, this step can be omitted.

```{r}
set.seed(1)
simpleCache("res2", {cmr(within, col_value = "sum", 
                         col_participant = "id",
                         col_dv = "dv", col_within = "iv_comb", nsample = 1e5)})
res2
res2$estimate
```

#### State-trace plot

Although the *stacmr* package provides a dedicated function for this purpose, it
does not produce a publication-ready plot. We therefore extract the relevant
data and construct the ST plot manually.

Note that in the thesis, standard errors were not calculated
correctly and were consequently overestimated. The approach used below follows
exactly the same procedure as implemented in the *stacmr* package.

```{r}
# Data for the Plot
d_agg <- data.frame(x_mean = stats$LibSum$means, y_mean = stats$IntSum$means,
                    x_se = sqrt(diag(stats$LibSum$cov) / diag(stats$LibSum$n)),
                    y_se = sqrt(diag(stats$IntSum$cov) / diag(stats$IntSum$n)),
                    iv_comb = names(stats$IntSum$conditions))

d_agg <- d_agg %>% mutate(x_min = x_mean - x_se,  # Define lower and upper limits for error bars
  x_max = x_mean + x_se,
  y_min = y_mean - y_se,
  y_max = y_mean + y_se)
d_agg

d_agg$iv_comb <- rownames(d_agg)

d_agg <- d_agg %>%
  separate(iv_comb, into = c("Self-Efficacy", "Artistic Interest"))
# correct factor ordering
d_agg$`Self-Efficacy` <- forcats::fct_relevel(d_agg$`Self-Efficacy`, c("low", "high"))
d_agg$`Artistic Interest` <- forcats::fct_relevel(d_agg$`Artistic Interest`, c("low", "medium", "high"))

d_cmr <- as.data.frame(res2$estimate)

# ggplot
p2 <- ggplot(d_agg, aes(x_mean, y_mean)) +
  geom_point(data = d_cmr, aes(LibSum, IntSum), color = "grey60",
             shape = 5, size = 3) +  # Add predicted Points (Regression)
  geom_line(data = d_cmr, aes(LibSum, IntSum), color = "grey60", linetype = "dashed") +  # grey curve for Regression model
  geom_errorbar(aes(ymin = y_min, ymax = y_max), color = "grey60", width = 0.2) +  # y-Axis Error bars
  geom_errorbar(aes(xmin = x_min, xmax = x_max), color = "grey60",
                orientation = "y", width = 0.2) +  # Add x-Axis Error Bars
  geom_point(aes(shape = `Artistic Interest`, fill = `Self-Efficacy`),
             color = "black", size = 3) +  # open low SE, filled high SE
  scale_shape_manual(values = c(21, 24, 22)) +  # fillable circle, triangle, square
  scale_fill_manual(values = c(low = "white", high = "black")) +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  labs(
    x = "Liberalism",
    y = "Intellect",
  ) +
   xlim(7, 17.5) +
   ylim(7, 17.5) +
  
  theme_bw() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.grid = element_blank(),
    legend.background = element_blank(),
    legend.key = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.box = "horizontal",
    legend.position = c(0.02, 0.98),   # x, y in NPC coordinates (0 = left/bottom, 1 = right/top)
    legend.justification = c("left", "top")
  ) +
  coord_fixed(ratio = 1) + 
  annotate("text", 11.4, 12.25, label = "dashed line shows\nbest isotonic model",
           hjust = 0)
p2
```
